SUPPORTING INFORMATION:\\ Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-09-11 Manuscript/manuscriptPSsuppl.tex

\section{Headgroup response to the additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig.~\ref{PSresponseTONaCl}). 
Experimental order parameters of POPC headgroup in the mixture are available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
in increasing NaCl concentration. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
The different binding behaviour is evident based on the response of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig.~\ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak K$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}.

In simulations with Berger and CHARMM36 models, the responses of POPC and POPS order parameter
to added sodium and potassium are not in line with the experiments. Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig.~\ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium 
in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
(Fig.~\ref{CIdensPSOCmixt}), but significantly larger error bars and
less systematic changes in the order parameters (Fig.~\ref{PSresponseTONaCl}).
Similar unsystematic behaviour was also observed in the simulations of Lipid14/17 model
with the additional counterions \cite{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa},
for which the data is not shown due to the formation of
ion clusters in water with relatively low (1~M) ion concentrations (Fig.~\ref{ionCLUSTERS}).
Appearance of such clusters in the MagRog simulations with 4~M KCl
could explain the unsystematic changes of the order parameters in this model upon increasing KCl.
In conclusion, the results are in line with the Section \ref{ciBINDINGsection}
in the main text, suggesting that the MacRog simulations with KCl give the most
realistic surface charge at the lipid bilayer interface among the tested simulation models.
\begin{figure*}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref.~\citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is shifted with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/lipid17cluster.png}
  \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1~M of NaCl (left)
    and MacRog simulations with 4~M of KCl (right).
\label{ionCLUSTERS}
  }
\end{figure}
\newpage

>>>>

\section{Headgroup response to additional counterions in POPC:POPS (5:1) mixtures}\label{mixtureTOadditionalCIs}
To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
Fig.~\ref{PSresponseTONaCl} plots the headgroup order parameters $S^\beta_\mathrm{CH}$ and $S^\alpha_\mathrm{CH}$ measured from a POPC:POPS (5:1) mixture
as a function of different monovalent ions added to the buffer. 
Experimental order parameters of the POPC headgroup in the mixture are available as a function
of LiCl and KCl concentrations, while the POPS headgroup order parameters are measured also
with increasing NaCl concentration. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
The different binding behavior is evident in the responses of the PS headgroup order parameters, as for example
%$S^\beta_\mathrm{CH}$ increases with Li$^+$, but decreases with K$^+$, Na$^+$ grows initially and then decreases;
%the smaller $S^\alpha_\mathrm{CH}$ decreases with all three monovalent cations;
the larger $S^\alpha_\mathrm{CH}$
decreases with Li$^+$, but increases with K$^+$ and Na$^+$, see Fig.~\ref{PSresponseTONaCl}.
The PC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating significant 
Li$^+$ binding, but only weak K$^+$ binding, when interpreted using the
molecular electrometer concept~\cite{akutsu81,altenbach84,seelig87}.

In simulations with Berger and CHARMM36, the responses of POPC and POPS $S_\mathrm{CH}$
to added sodium and potassium are not in line with the experiments. Rather, the simulations produce responses similar to experiments conducted in LiCl (Fig.~\ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium 
in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
(Fig.~\ref{CIdensPSOCmixt}), but significantly larger error bars and
less systematic changes in the order parameters (Fig.~\ref{PSresponseTONaCl}).
Similar unsystematic behavior was also observed in the simulations of Lipid14/17 model
with the additional counterions \cite{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa},
for which the data are not shown due to the formation of
ion clusters in water at relatively low (1~M) ion concentrations (Fig.~\ref{ionCLUSTERS}).
Appearance of such clusters in the MagRog simulations at 4~M KCl
could explain the unsystematic changes of the order parameters in this model upon increasing KCl.
In conclusion, the results are in line with the section 'Counterion binding and interactions between PC and PS headgroups' % \ref{ciBINDINGsection}
in the main text, suggesting that the MacRog simulations with KCl gave the most
realistic surface charge at the POPC:POPS (5:1) lipid bilayer interface among the tested simulation models.
\begin{figure*}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref.~\citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in the PS headgroup (bottom right), the $y$-axis is shifted with the same value for both $S^\alpha_\mathrm{CH}$ such that the lower order
    parameter value is at zero.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/lipid17cluster.png}
  \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
  \caption{Ion clusters appearing in POPC:POPS (5:1) Lipid14/17 simulations with 1~M of NaCl (left)
    and MacRog simulations with 4~M of KCl (right).
\label{ionCLUSTERS}
  }
\end{figure}
\newpage
